---
title: "Lab 2"
author: "Afrah Boateng, Mackenzie Henderson, Olivia Jackson Lambert, Tyler Kero" 
date: today 
date-format: long
format: 
  pdf: 
    documentclass: scrartcl
    classoption: onecolumn
    fontsize: 11pt
editor: visual
geometry: "left=1in,right=1in,top=1in,bottom=1in"
---

```{r load packages, include = FALSE}
#install.packages("ggplot2")
#install.packages("rlang")
#install.packages("quarto")
#install.packages("patchwork")
library(patchwork)
library(ggplot2)
library(tidyverse)
library(dplyr)
library(stargazer)
library(sandwich)
library(tidygeocoder)
library(geosphere)
```

```{r load data, include = FALSE}
games <- read.csv("../data/raw/Games.csv")
```

```{r old_code, include = FALSE}
games <- games %>%
  mutate(point_diff = awayScore - homeScore)

#There are 45 unique hometeamCity
unique(games$hometeamCity)

#There are 49 unique awayteamCity
unique(games$awayteamCity)

# Find the total Number of Unique Cities in the Dataset
unique_cities <- games %>%
  select(hometeamCity, awayteamCity) %>%         
  pivot_longer(cols = everything(),              
               names_to = "type",                
               values_to = "city") %>% 
  distinct(city)                                 

# View the unique cities
print(unique_cities)

# Count how many unique cities there are
num_unique_cities <- nrow(unique_cities)
print(num_unique_cities)

##There are 51 total unique cities (home and away)

# Clean city names by creating new city variable
city_lookup <- games %>%
  select(hometeamCity) %>% rename(city = hometeamCity) %>%
  bind_rows(games %>% select(awayteamCity) %>% rename(city = awayteamCity)) %>%
  distinct() %>%
  mutate(city_clean = case_when(
    city == "Los Angeles"            ~ "Los Angeles, California",
    city == "LA"                      ~ "Los Angeles, California",
    city == "Phoenix"                 ~ "Phoenix, Arizona",
    city == "New York"                ~ "New York, New York",
    city == "Cleveland"               ~ "Cleveland, Ohio",
    city == "Toronto"                 ~ "Toronto, Ontario, Canada",
    city == "Charlotte"               ~ "Charlotte, North Carolina",
    city == "Brooklyn"                ~ "Brooklyn, New York",
    city == "Oklahoma City"           ~ "Oklahoma City, Oklahoma",
    city == "Detroit"                 ~ "Detroit, Michigan",
    city == "Boston"                  ~ "Boston, Massachusetts",
    city == "Indiana"                 ~ "Indianapolis, Indiana",
    city == "Washington"              ~ "Washington, D.C.",
    city == "Milwaukee"               ~ "Milwaukee, Wisconsin",
    city == "Portland"                ~ "Portland, Oregon",
    city == "Memphis"                 ~ "Memphis, Tennessee",
    city == "Chicago"                 ~ "Chicago, Illinois",
    city == "Philadelphia"            ~ "Philadelphia, Pennsylvania",
    city == "San Antonio"             ~ "San Antonio, Texas",
    city == "Minnesota"               ~ "Minneapolis, Minnesota",
    city == "Denver"                  ~ "Denver, Colorado",
    city == "Utah"                    ~ "Salt Lake City, Utah",
    city == "Dallas"                  ~ "Dallas, Texas",
    city == "Golden State"            ~ "San Francisco, California",
    city == "Miami"                   ~ "Miami, Florida",
    city == "Houston"                 ~ "Houston, Texas",
    city == "New Orleans"             ~ "New Orleans, Louisiana",
    city == "Sacramento"              ~ "Sacramento, California",
    city == "Atlanta"                 ~ "Atlanta, Georgia",
    city == "Orlando"                 ~ "Orlando, Florida",
    city == "New Jersey"              ~ "New Jersey, USA",
    city == "Seattle"                 ~ "Seattle, Washington",
    city == "Vancouver"               ~ "Vancouver, British Columbia, Canada",
    city == "Kansas City"             ~ "Kansas City, Missouri",
    city == "San Diego"               ~ "San Diego, California",
    city == "Buffalo"                 ~ "Buffalo, New York",
    city == "Kansas City-Omaha"       ~ "Kansas City / Omaha, USA",
    city == "Capital"                 ~ "Washington, D.C.",
    city == "Baltimore"               ~ "Baltimore, Maryland",
    city == "Cincinnati"              ~ "Cincinnati, Ohio",
    city == "San Francisco"            ~ "San Francisco, California",
    city == "St. Louis"               ~ "St. Louis, Missouri",
    city == "Syracuse"                ~ "Syracuse, New York",
    city == "Minneapolis"             ~ "Minneapolis, Minnesota",
    city == "Ft. Wayne Zollner"       ~ "Fort Wayne, Indiana",
    city == "Rochester"               ~ "Rochester, New York",
    city == "Tri-Cities"              ~ "Tri-Cities, Washington / Oregon",
    city == "Guangzhou"               ~ "Guangzhou, Guangdong, China",
    city == "South East Melbourne"    ~ "Melbourne, Victoria, Australia",
    city == "Hapoel"                  ~ "Tel Aviv, Israel",
    TRUE                              ~ city
  ))

# Geocode to get lat/long
city_coords <- city_lookup %>%
  geocode(city_clean, method = "osm")   # or “iq” or “google” if you have API
# This will return lat & long columns

# Inspect:
head(city_coords)

#merge home team city coords to assign lat lon to each game
games2 <- games %>%
  left_join(city_coords %>% select(city, lat_home = lat, lon_home = long),
            by = c("hometeamCity" = "city")) %>%
  # Then merge away team city coords
  left_join(city_coords %>% select(city, lat_away = lat, lon_away = long),
            by = c("awayteamCity" = "city"))

#Count NAs in each column for games2
na_counts <- games2 %>%
  summarise(
    na_lat_home = sum(is.na(lat_home)),
    na_lon_home = sum(is.na(lon_home)),
    na_lat_away = sum(is.na(lat_away)),
    na_lon_away = sum(is.na(lon_away))
  )

print(na_counts)

#Find which rows have any NAs in those columns for games2
rows_with_na <- games2 %>%
  filter(
    is.na(lat_home) | is.na(lon_home) | is.na(lat_away) | is.na(lon_away)
  ) %>%
  select(hometeamCity, awayteamCity, lat_home, lon_home, lat_away, lon_away)


## 71 Rows with NA lat lon, all due to Tri-cities, which i don't think even exists anymore.
## We can ignore these; they will be dropped from the regression automatically

games3 <- games2 %>%
  mutate(
    dist_meters = distHaversine(cbind(lon_home, lat_home),
                                cbind(lon_away, lat_away)),
    dist_miles = dist_meters * 0.000621371,
    year = as.numeric(substr(gameDate, 1, 4)),
    winning_team = case_when(
      winner == awayteamId ~ "away",
      winner == hometeamId ~ "home",
      TRUE ~ NA_character_
    ),
    
    # ------------ Longitude-based zones ----------
    home_zone = case_when(
      lon_home < -105 ~ "West",
      lon_home > -90  ~ "East",
      TRUE            ~ "Central"   # in between → Central
    ),
    away_zone = case_when(
      lon_away < -105 ~ "West",
      lon_away > -90  ~ "East",
      TRUE            ~ "Central"
    )
  )


games4 <- games3 %>%
  filter(gameType == "Regular Season", !is.na(dist_miles)) %>%
  mutate(
    away_travel = case_when(
      away_zone == "East"    & home_zone == "West"    ~ "East to West",
      away_zone == "West"    & home_zone == "East"    ~ "West to East",
      away_zone == "East"    & home_zone == "Central" ~ "East to Central",
      away_zone == "Central" & home_zone == "East"    ~ "Central to East",
      away_zone == "Central" & home_zone == "West"    ~ "Central to West",
      away_zone == "West"    & home_zone == "Central" ~ "West to Central",
      away_zone == home_zone                         ~ "In-Zone",
      TRUE                                          ~ "Other/International"
    ),
    away_travel_high = factor(
      case_when(
        away_travel == "East to West" ~ "East to West",
        away_travel == "West to East" ~ "West to East",
        TRUE                          ~ "Neither"
      ),
      levels = c("Neither", "West to East", "East to West")
    )
  )

set.seed(123)   # ensures reproducibility

# total number of rows
n <- nrow(games4)

# sample 30% of the rows for EDA
eda_index <- sample(1:n, size = floor(0.30 * n), replace = FALSE)

# create the two datasets
games4_eda   <- games4[eda_index, ]
games4_model <- games4[-eda_index, ]


lm_model_1 <- lm(point_diff ~ dist_miles, data = games4_model)

lm_model_2 <- lm(point_diff ~ dist_miles + away_travel_high, data = games4_model)

lm_model_3 <- lm(point_diff ~ dist_miles + away_travel_high + attendance, data = games4_model)

lm_model_4 <- lm(point_diff ~ dist_miles + away_travel_high + attendance + year, data = games4_model)
```

\newpage

# **Introduction**

Professional sports leagues such as the NBA require teams to travel frequently across large geographic distances, often across multiple time zones, within tightly scheduled seasons. Travel has long been discussed as a potential contributor to competitive imbalance, fatigue, and home-court advantage, yet the extent to which travel-related factors are reflected in game outcomes remains unclear. Descriptively characterizing how performance varies with travel conditions provides insight into the structure of competition in the league, independent of any attempt to predict outcomes or identify causal mechanisms.\

We propose to examine the relationship between travel distance and game performance in the NBA, specifically whether the number of miles an away team travels affects point differential (away team score minus home team score). This research question has practical implications for multiple stakeholders: coaches and team administrators can use these insights for scheduling and travel logistics; players and athletic trainers can better prepare for the physical demands of travel; and sports bettors can make more informed predictions. While previous research, notably Deddens and Steenland (1997),  found few consistent effects of travel distance in their analysis of regular season NBA games, we aim to revisit this question with a more comprehensive dataset spanning seven decades. Additionally, we will explore related questions such as whether travel direction or game attendance levels moderate the relationship between distance travelled and performance.

# Data

### Data Source

The dataset was sourced from Kaggle’s “NBA Dataset - Box Scores & Stats”. This dataset contains data on NBA games from 1947 to 2025 and is updated regularly with official statistics from the NBA’s website. It contains 72,278 observations, with a unit of observation being a single game, and 17 variables such as game attendance and points for each team.

### **Data Wrangling**

The dataset was restricted to regular season games, reducing the size by 10%. 71 games were removed where the city was indeterminate, such as the historic Tri-State team, a midwest team from the early years of the NBA that was not tied to a specific city. For each unit of observation, seven additional variables were calculated: (1) the point differential between the away and home teams, (2) home team latitude, (3) home team longitude, (4) away team latitude, (5) away team longitude, (6) the geodesic distance between the home and away team cities, and finally (7) a categorical variable describing whether away teams were travelling from the East Coast to the West Coast. 

This categorical variable contains 3 possibilities: travel from east to west coast, travel from west to east coast or neither, which encompasses all travel within the same region or travel between the central regions of the US and coasts. The point differential was calculated by subtracting the home team’s final points from the away team’s final points, so that a positive point differential translates to a win for the away team. Lastly, the latitude and longitude of cities were calculated using geocodes from OpenStreetMap based on the cleaned city name for each team. 

After wrangling, the final dataset consisted of 65,467 games, of which 30% (19,640) were randomly selected for exploratory data analysis and 70% (45,827) were set aside for modeling.

### **Operationalization**

To capture the relationship between away team travel and game outcomes, the following variables were included in our analysis:

*Output Variable*

-   Point Differential: Captures the outcome of a game and the magnitude of a team’s win.

*Input Variables*

-   Distance: The geodesic distance travelled by the away team.

-   Travel Direction: Tracks the direction of movement and if it was from coast-to-coast.

-   Attendance: Game attendance, which may affect the performance of each team.

-   Year: The year in which the game was played. Included due to findings in our initial exploration which suggested a strong decrease in the percentage of home team wins over time.

```{r example_cell, echo=FALSE, warning=FALSE, message=FALSE, fig.width = 7, fig.height=2.6}

```

**Figure 1:** Number of respondents identifying with each party description.

# **Modeling**

### **Model Specification**

Large sample linear model

```{r linear_regression, echo = FALSE, results='asis', message=FALSE, warning=FALSE}

stargazer(
  lm_model_1, lm_model_2, lm_model_3, lm_model_4,
  header = FALSE,
  no.space = TRUE, 
  column.sep.width = "0.5pt", 
  omit.stat = "f",
  font.size = "footnotesize",
  type="latex",
  title="Linear Regression Co-efficients for Point Differential Against Input Variables", 
  dep.var.labels = "Point Differential", 
  covariate.labels = c("Distance Travelled (miles)", 
                       "West to East Direction", 
                       "East to West Direction",
                       "Game Attendance",
                       "Year",
                       "Constant"), 
  digits=3)
```

\

### **Model Assumptions**

#### Independent & Identically Distributed

#### A Unique BLP Exists

### **Model Results & Interpretation**

### **Scope & Limitations**

# **Conclusion**

\newpage

# Citations

# Appendix
