---
title: "Lab 2"
author: "Afrah Boateng, Mackenzie Henderson, Olivia Jackson Lambert, Tyler Kero" 
date: today 
date-format: long
format: 
  pdf: 
    documentclass: scrartcl
    classoption: onecolumn
    fontsize: 11pt
editor: visual
geometry: "left=2cm,right=2cm,top=2cm,bottom=2.5cm"
---

```{r load_packages, include = FALSE}
library(patchwork)
library(ggplot2)
library(tidyverse)
library(dplyr)
library(stargazer)
library(sandwich)
library(tidygeocoder)
library(geosphere)
theme_set(theme_minimal())
```

```{r load_data, include = FALSE}
games <- readRDS("../data/processed/games_clean_model.rds")
```

\newpage

## **Introduction**

Professional sports leagues such as the NBA require teams to travel frequently across large geographic distances, often across multiple time zones, within tightly scheduled seasons. Travel has long been discussed as a potential contributor to competitive imbalance, fatigue, and home-court advantage, yet the extent to which travel-related factors are reflected in game outcomes remains unclear. Understanding these patterns can help coaches optimize scheduling and travel logistics, assist players and athletic trainers in preparing for physical demands, and inform sports bettors’ predictions.

We propose to examine whether the distance an away team travels affects point differential (away team score minus home team score). Our work builds upon previous research, including Deddens and Steenland (1997),  which found limited effects of travel distance in their analysis of regular season NBA games. We aim to revisit this question by leveraging a comprehensive dataset spanning across seven decades to provide a more complete picture of travel-related performance patterns.

## Data

*Data Source -* The data set was sourced from Kaggle’s “NBA Dataset - Box Scores & Stats”. This data set contains data on NBA games from 1947 to 2025 and is updated regularly with official statistics from the NBA’s website. It contains 72,278 games and 17 variables such as game attendance and points for each team.

*Data Wrangling -* The data set was restricted to regular season games, reducing the size by 10%. 71 games were removed where the city was indeterminate, such as the historic mid-west Tri-Cities team. Games with travel distances larger than 3500 miles or which had attendance above 100,000 were filtered from the dataset to focus on US travel and remove anomalies. For each unit of observation, seven additional variables were calculated: (1) the point differential between the away and home teams, (2) home team latitude, (3) home team longitude, (4) away team latitude, (5) away team longitude, (6) the geodesic distance between the home and away team cities, and finally (7) a categorical variable indicating coast to coast travel for the away team.

This categorical variable contains 3 possibilities: travel from east to west coast, travel from west to east coast, or neither, which encompasses all travel within the same region or travel between the central regions of the US and coasts. The point differential was calculated by subtracting the home team’s score from the away team’s score, so that a positive point differential translates to a win for the away team. Lastly, the latitude and longitude of cities were calculated using geocodes from OpenStreetMap based on the cleaned city name for each team.

After wrangling, the final data set consisted of 67,004 games, of which 30% (20,101) were randomly selected for exploratory data analysis and 70% (46,903) were set aside for modeling.

*Operationalization -* To capture the relationship between away team travel and game outcomes, the following variables were included in our analysis:

-   Point Differential (Dependent): Captures the direction and magnitude of the game outcome from the away team’s perspective.
-   Distance: The geodesic distance traveled by the away team.
-   Travel Direction: Tracks the direction of movement and if it was from coast-to-coast.
-   Attendance: Game attendance, which may affect the comfort of each team and thereby tehir performances.
-   Year: The year in which the game was played. Included due to findings in our initial exploration which suggested a strong decrease in the percentage of home team wins over time.

```{r eda_graphs, echo=FALSE, warning=FALSE, message=FALSE, fig.width = 9, fig.height=6}

games_eda <- readRDS("../data/processed/games_clean_eda.rds")

a1 <- ggplot(games_eda, aes(x = dist_miles, y = point_diff)) +
  geom_point(alpha = 0.5, size = 0.5, color = "royalblue4") +
  geom_smooth(method = "lm", color = "red", se = TRUE) +
  labs(
    x = "Distance Travelled (miles)",
    y = "Point Differential",
    title = "Point Differential vs. Distance Travelled"
  ) +
  theme(
  plot.title = element_text(size = 10, hjust = 0.5, face = "bold"),
  axis.title = element_text(size = 9),
  axis.text = element_text(size = 8)
  )

a2 <- ggplot(games_eda, aes(x = year, y = point_diff)) +
  geom_point(alpha = 0.5, size = 0.5, color = "royalblue4") +
  geom_smooth(method = "lm", color = "red", se = TRUE) +
  labs(
    x = "Year",
    y = "Point Differential",
    title = "Point Differential vs. Year"
  ) +
  theme(
  plot.title = element_text(size = 10, hjust = 0.5, face = "bold"),
  axis.title = element_text(size = 9),
  axis.text = element_text(size = 8)
  )

a3 <- ggplot(games_eda, aes(x = attendance, y = point_diff)) +
  geom_point(alpha = 0.5, size = 0.5, color = "royalblue4") +  # tiny, semi-transparent points
  geom_smooth(method = "lm", color = "red", se = TRUE) +      # regression line with confidence interval
  labs(
    x = "Attendance",
    y = "Point Differential",
    title = "Point Differential vs. Attendance"
  ) +
  theme(
  plot.title = element_text(size = 10, hjust = 0.5, face = "bold"),
  axis.title = element_text(size = 9),
  axis.text = element_text(size = 8)
  )

a4 <- ggplot(games_eda, aes(x = away_travel_high, y = point_diff)) +
  geom_boxplot(fill = "snow", color = "royalblue4") +  # Boxplot with color
  labs(
    x = "Travel Direction",
    y = "Point Differential",
    title = "Point Differential vs. Travel Direction"
  ) +
  theme(
  plot.title = element_text(size = 10, hjust = 0.5, face = "bold"),
  axis.title = element_text(size = 9),
  axis.text = element_text(size = 8)
  )

a1 + a2 + a3 + a4 + plot_layout(ncol = 2)
```

**Figure 1:** Exploratory data analysis of point differential and independent variables.

Before modeling, the relationships between the input and response variables were explored. From initial exploration, the scatter plot of point differential vs. travel distance shows no clear trend, with much noise around zero. This suggests a weak descriptive relationship between the two variables, and it is likely that other factors play a much larger role in determining game outcomes.

The scatter plot of point differential and year show that the range of the point differentials has gradually increased over time, which can be seen from the fan shape. This suggests that the distribution of game outcomes has widened in more recent years, potentially reflecting structural changes in the NBA which have led to more pronounced wins and losses.

For point differential and attendance, the spread of point differentials increases with attendance up to approximately 20,000 spectators. This pattern suggests greater variability in game outcomes among games with higher attendance.

Finally, the box plots comparing point differential across the three travel direction categories show that west to east travel is associated with a slightly higher median point differential for away teams compared to other categories. This suggests that travel direction may be meaningfully related to away-team performance, motivating its inclusion in the analysis.

## **Modeling**

```{r linear_regression_models, include = FALSE}

lm_model_1 <- lm(point_diff ~ dist_miles, data = games)

lm_model_2 <- lm(point_diff ~ dist_miles + away_travel_high, data = games)

lm_model_3 <- lm(point_diff ~ dist_miles + away_travel_high + attendance, data = games)

lm_model_4 <- lm(point_diff ~ dist_miles + away_travel_high + attendance + year, data = games)
```

### **Model Specification**

Given the large sample size, we use a linear regression model to examine the relationship between point differential and variables such as travel distance, travel direction, game attendance, and year. The regression table below shows four models that progressively add more predictors.

*Model 1*: The first model includes only travel distance. The coefficient for this variable is not statistically significant, suggesting that travel distance has little effect on point differential.

*Model 2*: In this model, we add travel direction (with "Neither" as the reference variable). The coefficients for West-to-East and East-to-West are statistically significant, indicating that teams traveling from the West Coast to the East Coast tend to perform better compared to those traveling from the East Coast to the West Coast.

*Model 3*: Game attendance is introduced. Although it is statistically significant, the effect is small, suggesting that larger crowds slightly affect the point differential but with limited practical significance..

*Model 4*: Finally, we add year. We observe a statistically significant positive effect, implying that point differentials have increased over time.

```{r model, echo = FALSE, results='asis', message=FALSE, warning=FALSE}

cov_1 <- vcovHC(lm_model_1, type = "HC1")
robust.se_1 <- sqrt(diag(cov_1))
robust.se_1 <- robust.se_1[names(coef(lm_model_1))]

cov_2 <- vcovHC(lm_model_2, type = "HC1")
robust.se_2 <- sqrt(diag(cov_2))
robust.se_2 <- robust.se_2[names(coef(lm_model_2))]

cov_3 <- vcovHC(lm_model_3, type = "HC1")
robust.se_3 <- sqrt(diag(cov_3))
robust.se_3 <- robust.se_3[names(coef(lm_model_3))]

cov_4 <- vcovHC(lm_model_4, type = "HC1")
robust.se_4 <- sqrt(diag(cov_4))
robust.se_4 <- robust.se_4[names(coef(lm_model_4))]


stargazer(
  lm_model_1, lm_model_2, lm_model_3, lm_model_4,
  se = list(
    robust.se_1,
    robust.se_2,
    robust.se_3,
    robust.se_4
  ),
  header = FALSE,
  no.space = TRUE, 
  column.sep.width = "0.5pt", 
  omit.stat = "f",
  font.size = "footnotesize",
  type="latex",
  title="Linear Regression Co-efficients for Point Differential Against Input Variables", 
  dep.var.labels = "Point Differential", 
  covariate.labels = c("Distance Travelled (miles)", 
                       "West to East Direction", 
                       "East to West Direction",
                       "Game Attendance",
                       "Year",
                       "Constant"), 
  digits=3)
```

### **Model Assumptions**

Since the model is a large sample linear model, the required assumptions are as follows:

*Independent & Identically Distributed Errors -* The linear regression framework assumes that error terms are independent and identically distributed. In this context, games are not fully independent, as team performance may be correlated across games due to persistent team quality, opponent strength, injuries, and broader league dynamics. While including year as a numeric covariate captures long-run temporal trends in point differentials, it does not fully eliminate within-team or within-season dependence. As a result, the model should be interpreted descriptively, with coefficient estimates reflecting average associations rather than causal effects.

*A Unique Best Linear Predictor Exists -* Covariance and correlation matrices (Appendix Figures X and Y) indicate finite variance and no perfect multicollinearity among predictors, satisfying the technical requirements for OLS estimation. However, some predictors, particularly travel distance and travel direction, are moderately to strongly correlated, reflecting the structural relationship between long-distance travel and coast-to-coast movement. As a result, individual coefficient estimates should be interpreted cautiously, as shared variation across predictors may affect their magnitude and statistical significance.

*A Note on Linear Relationships -* The model assumes linear relationships between point differential and continuous predictors such as distance, attendance, and year. Exploratory data analysis indicates substantial variability around an approximately linear trend, which may mask nonlinear patterns, particularly across a seven-decade sample that encompasses significant structural changes in the NBA. As a result, the estimated coefficients should be interpreted as approximations to average linear relationships rather than exact functional forms.

## **Model Results**

*Interpretation -* The final model (Model 4) is statistically significant as indicated by the F-statistic (p \< 2.2e−16). However, the overall explanatory power of the model remains very limited, with an R² of 0.005. Given the large sample size, small effects may achieve statistical significance without corresponding practical importance. As a result, emphasis is placed on the magnitude and direction of estimated effects rather than statistical significance alone. The coefficient on distance traveled is −0.0002, implying that, holding other variables constant, each additional mile traveled by the away team is associated with a 0.0002-point decrease in point differential. While the direction of the effect aligns with intuition, the estimate is neither statistically significant nor practically meaningful. For example, even a 3,000-mile trip corresponds to an expected decrease of only 0.6 points, which is small relative to typical game-to-game variation.

In contrast, travel direction exhibits statistically significant effects, with West-to-East travel associated with higher point differentials for away teams, suggesting that time-zone alignment may matter more than simply distance traveled. Game attendance exhibits a small but statistically significant negative association with point differential. This result reflects a shift in the mean point differential rather than changes in its dispersion. While exploratory analysis suggested greater variability in point differentials at higher attendance levels, the regression estimates indicate that, on average, higher attendance is associated with slightly smaller away-team advantages. The effect size is small and should be interpreted cautiously.

Overall, the results suggest that while geographic orientation and temporal trends influence competitive outcomes, travel distance alone is not a primary driver of game performance.

*Scope & Limitations -* This analysis is descriptive in nature and does not support causal claims about the effect of travel on NBA game outcomes. Games are not fully independent, as team performance may be correlated across time due to persistent team quality, injuries, and broader league dynamics. Additionally, while geodesic distance captures the physical separation between teams, it does not directly measure fatigue, rest, travel conditions, or circadian disruption, such as back-to-back games or sleep loss from changing time zones. Finally, structural changes in the NBA over the seven-decade sample period, including standardization of courts, more movement of players between teams, and traveling being less strenuous, may influence both the distribution of point differentials and their relationship with travel-related variables.

## **Conclusion**

Our analysis finds little evidence that travel distance alone is meaningfully associated with point differential in NBA regular season games. While the estimated effect of distance is negative, it is small in magnitude and not statistically distinguishable from zero. In contrast, travel direction exhibits more consistent associations, with West-to-East travel linked to higher away-team point differentials, suggesting that geographic position and time-zone alignment may play a larger role than distance itself.

\newpage

## Citations

Effect of Travel and Rest on Performance of Professional Basketball Players. (1997). *Sleep*. https://doi.org/10.1093/sleep/20.5.366

## Appendix

Link to the Dataset : <https://www.kaggle.com/datasets/eoinamoore/historical-nba-data-and-player-box-scores>

```{r test_graphs, echo = FALSE, message=FALSE, warning=FALSE, fig.width = 9, fig.height=9}
lm_model_4 <- lm(point_diff ~ dist_miles + away_travel_high + attendance + year, data = games)
games$fitted_values <- predict(lm_model_4, newdata = games)

p1 <- ggplot(games, aes(x = point_diff, y = fitted_values)) +
  geom_point(alpha = 0.5, size = 1, color = "royalblue4") +  # Raw data points
  geom_smooth(method = "lm", color = "red", se = FALSE) +  # Regression line
  labs(
    title = "Actual Point Difference vs. Distance Travelled",
    x = "Distance Travelled (miles)",
    y = "Actual Point Difference"
  ) +
  theme(
  plot.title = element_text(size = 10, hjust = 0.5, face = "bold"),
  axis.title = element_text(size = 9),
  axis.text = element_text(size = 8)
  )

p2 <- ggplot(games, aes(x = dist_miles, y = fitted_values)) +
  geom_point(alpha = 0.5, size = 1, color = "royalblue4") +  # Raw data points
  geom_smooth(method = "lm", color = "red", se = FALSE) +  # Regression line
  labs(
    title = "Modelled Point Difference vs. Distance Travelled",
    x = "Distance Travelled (miles)",
    y = "Predicted Point Difference"
  ) +
  theme(
  plot.title = element_text(size = 10, hjust = 0.5, face = "bold"),
  axis.title = element_text(size = 9),
  axis.text = element_text(size = 8)
  )

p3 <- ggplot(games, aes(x = year, y = fitted_values)) +
  geom_point(alpha = 0.5, size = 1, color = "royalblue4") +  # Raw data points
  geom_smooth(method = "lm", color = "red", se = FALSE) +  # Regression line
  labs(
    title = "Modelled Point Difference vs. Year",
    x = "Year",
    y = "Predicted Point Difference"
  ) +
  theme(
  plot.title = element_text(size = 10, hjust = 0.5, face = "bold"),
  axis.title = element_text(size = 9),
  axis.text = element_text(size = 8)
  )

p4 <- ggplot(games, aes(x = attendance, y = fitted_values)) +
  geom_point(alpha = 0.5, size = 1, color = "royalblue4") +  # Raw data points
  geom_smooth(method = "lm", color = "red", se = FALSE) +  # Regression line
  labs(
    title = "Modelled Point Difference vs. Attendance",
    x = "Attendance",
    y = "Predicted Point Difference"
  ) +
  theme(
  plot.title = element_text(size = 10, hjust = 0.5, face = "bold"),
  axis.title = element_text(size = 9),
  axis.text = element_text(size = 8)
  )

p5 <- ggplot(games, aes(x = away_travel_high, y = fitted_values)) +
  geom_boxplot(fill = "snow", color = "royalblue4") +
  labs(title = "Modelled Point Difference for Travel Direction", x = "Away Travel High", y = "Predicted Point Difference") +
  theme(
  plot.title = element_text(size = 10, hjust = 0.5, face = "bold"),
  axis.title = element_text(size = 9),
  axis.text = element_text(size = 8)
  )

games$fitted_values <- predict(lm_model_4, newdata = games)
games$residuals <- games$point_diff - games$fitted_values

p6 <- ggplot(games, aes(x = fitted_values, y = residuals)) +
  geom_point(alpha = 0.5, size = 1, color = "royalblue4") +  # Raw data points
  geom_hline(yintercept = 0, color = "red") +  # Add a horizontal line at y = 0
  labs(
    title = "Residuals vs Predicted Values",
    x = "Predicted Values (Fitted)",
    y = "Residuals"
  ) +
  theme(
  plot.title = element_text(size = 10, hjust = 0.5, face = "bold"),
  axis.title = element_text(size = 9),
  axis.text = element_text(size = 8)
  )

p1+p2+p3+p4+p5+p6 + plot_layout(ncol = 2)
```

```{r cov_data_clean, echo = FALSE, results='asis', message=FALSE, warning=FALSE, fig.width=150}
cov_cor_cleaned_data <- na.omit(games[, c("dist_miles", "year", "attendance", "away_travel_high", "point_diff")])
cov_cor_cleaned_data$travel_direction <- as.numeric(cov_cor_cleaned_data$away_travel_high)
```

```{r covariance_matrix, echo = FALSE, results='asis', message=FALSE, warning=FALSE, fig.width=150}
library(knitr)

cov_mat <- cov(cov_cor_cleaned_data[, c(
  "dist_miles", "year", "attendance", "travel_direction", "point_diff"
)])

# Round values for readability
cov_mat_rounded <- round(cov_mat, 2)

# Convert to data frame and keep variable names
cov_df <- as.data.frame(cov_mat_rounded) |> 
  rownames_to_column("Variable")

# Render nicely with a title
kable(
  cov_df,
  caption = "Covariance Matrix of Model Variables",
  booktabs = TRUE
)

```

```{r correlation_matrix,  echo = FALSE, results='asis', message=FALSE, warning=FALSE}

library(knitr)
library(tibble)

# Compute the correlation matrix
cor_mat <- cor(cov_cor_cleaned_data[, c(
  "dist_miles", "year", "attendance", "travel_direction", "point_diff"
)])

# Round values for readability
cor_mat_rounded <- round(cor_mat, 2)

# Convert to data frame and keep variable names
cor_df <- as.data.frame(cor_mat_rounded) |> 
  rownames_to_column("Variable")

# Render nicely with a title
kable(
  cor_df,
  caption = "Correlation Matrix of Model Variables",
  booktabs = TRUE
)
```
