---
title: "Lab 2"
author: "Afrah Boateng, Mackenzie Henderson, Olivia Jackson Lambert, Tyler Kero" 
date: today 
date-format: long
format: 
  pdf: 
    documentclass: scrartcl
    classoption: onecolumn
    fontsize: 11pt
editor: visual
geometry: "left=1in,right=1in,top=1in,bottom=1in"
---

```{r load_packages, include = FALSE}
library(patchwork)
library(ggplot2)
library(tidyverse)
library(dplyr)
library(stargazer)
library(sandwich)
library(tidygeocoder)
library(geosphere)

theme(
  plot.title = element_text(size = 10),
  axis.title = element_text(size = 9),
  axis.text = element_text(size = 8)
)

theme_set(theme_minimal())
```

```{r load_data, include = FALSE}
games <- readRDS("../data/processed/games_clean_model.rds")
```

\newpage

# **Introduction**

Professional sports leagues such as the NBA require teams to travel frequently across large geographic distances, often across multiple time zones, within tightly scheduled seasons. Travel has long been discussed as a potential contributor to competitive imbalance, fatigue, and home-court advantage, yet the extent to which travel-related factors are reflected in game outcomes remains unclear. Descriptively characterizing how performance varies with travel conditions provides insight into the structure of competition in the league, independent of any attempt to predict outcomes or identify causal mechanisms.\

We propose to examine the relationship between travel distance and game performance in the NBA, specifically whether the number of miles an away team travels affects point differential (away team score minus home team score). This research question has practical implications for multiple stakeholders: coaches and team administrators can use these insights for scheduling and travel logistics; players and athletic trainers can better prepare for the physical demands of travel; and sports bettors can make more informed predictions. While previous research, notably Deddens and Steenland (1997), found few consistent effects of travel distance in their analysis of regular season NBA games, we aim to revisit this question with a more comprehensive data set spanning seven decades. Additionally, we will explore related questions such as whether travel direction or game attendance levels moderate the relationship between distance traveled and performance.

# Data

*Data Source -* The data set was sourced from Kaggle’s “NBA Dataset - Box Scores & Stats”. This data set contains data on NBA games from 1947 to 2025 and is updated regularly with official statistics from the NBA’s website. It contains 72,278 observations, with a unit of observation being a single game, and 17 variables such as game attendance and points for each team.

*Data Wrangling -* The data set was restricted to regular season games, reducing the size by 10%. 71 games were removed where the city was indeterminate, such as the historic Tri-State team, a mid-west team from the early years of the NBA that was not tied to a specific city. For each unit of observation, seven additional variables were calculated: (1) the point differential between the away and home teams, (2) home team latitude, (3) home team longitude, (4) away team latitude, (5) away team longitude, (6) the geodesic distance between the home and away team cities, and finally (7) a categorical variable describing whether away teams were travelling from the East Coast to the West Coast. 

This categorical variable contains 3 possibilities: travel from east to west coast, travel from west to east coast or neither, which encompasses all travel within the same region or travel between the central regions of the US and coasts. The point differential was calculated by subtracting the home team’s final points from the away team’s final points, so that a positive point differential translates to a win for the away team. Lastly, the latitude and longitude of cities were calculated using geocodes from OpenStreetMap based on the cleaned city name for each team. 

After wrangling, the final data set consisted of 65,467 games, of which 30% (19,640) were randomly selected for exploratory data analysis and 70% (45,827) were set aside for modeling.

*Operationalization -* To capture the relationship between away team travel and game outcomes, the following variables were included in our analysis:

-   Point Differential (Dependent): Captures the outcome of a game and the magnitude of a team’s win.
-   Distance: The geodesic distance traveled by the away team.
-   Travel Direction: Tracks the direction of movement and if it was from coast-to-coast.
-   Attendance: Game attendance, which may affect the performance of each team.
-   Year: The year in which the game was played. Included due to findings in our initial exploration which suggested a strong decrease in the percentage of home team wins over time.

```{r eda_graphs, echo=FALSE, warning=FALSE, message=FALSE, fig.width = 9, fig.height=6}

games_eda <- readRDS("../data/processed/games_clean_eda.rds")

a1 <- ggplot(games_eda, aes(x = dist_miles, y = point_diff)) +
  geom_point(alpha = 0.5, size = 0.5, color = "blue") +
  geom_smooth(method = "lm", color = "red", se = TRUE) +
  xlim(0, 3000) +
  labs(
    x = "Distance Travelled (miles)",
    y = "Point Differential",
    title = "Point Differential vs. Distance Travelled"
  )

a2 <- ggplot(games_eda, aes(x = year, y = point_diff)) +
  geom_point(alpha = 0.5, size = 0.5, color = "blue") +
  geom_smooth(method = "lm", color = "red", se = TRUE) +
  labs(
    x = "Year",
    y = "Point Differential",
    title = "Point Differential vs. Year"
  )

a3 <- ggplot(games_eda, aes(x = attendance, y = point_diff)) +
  geom_point(alpha = 0.5, size = 0.5, color = "blue") +  # tiny, semi-transparent points
  geom_smooth(method = "lm", color = "red", se = TRUE) +      # regression line with confidence interval
  xlim(0, 75000) +
  labs(
    x = "Attendance",
    y = "Point Differential",
    title = "Point Differential vs. Attendance"
  )

a4 <- ggplot(games_eda, aes(x = away_travel_high, y = point_diff)) +
  geom_boxplot(fill = "lightblue", color = "darkblue") +  # Boxplot with color
  labs(
    x = "Travel Direction",
    y = "Point Differential",
    title = "Point Differential vs. Travel Direction"
  )

a1 + a2 + a3 + a4 + plot_layout(ncol = 2)
```

**Figure 1:** Exploratory Data Analysis of Point Differential and Predictor Variables

To take a first pass at the data, we examine the relationships between each individual predictor variable (Distance Traveled, Year, Attendance, and Travel Direction) and the response variable (Point Differential). The scatter plot between Point Differential and Distance Traveled shows no clear trend, suggesting a minimal relationship between the two variables. This indicates that the distance teams travel doesn’t have a strong impact on the point differential, and other factors likely contribute more significantly to game outcomes. Next, the scatter plot between Point Differential and Year displays that the point differential range has been slowly increasing as the year has increased. This may indicate that the evolution of the NBA has led to more pronounced wins and losses over time, suggesting that disparities in team performance have become more significant as the years have passed. Next, the scatter plot between Point Differential and Attendance shows that games with a higher attendance often have a greater point differential spread, with games around 20,000 in attendance being more likely to feature a greater margin of victory or defeat than those with attendance ranging from 0 to 10,000. Finally, the box plots exhibiting Point Differential between the three travel direction buckets indicate that West to East travel has an average higher point differential than Neither or East to West. This suggests that teams traveling from West to East may face additional challenges that impact their performance, possibly due to factors such as time zone changes or fatigue.

# **Modeling**

```{r linear_regression_models, include = FALSE}

lm_model_1 <- lm(point_diff ~ dist_miles, data = games)

lm_model_2 <- lm(point_diff ~ dist_miles + away_travel_high, data = games)

lm_model_3 <- lm(point_diff ~ dist_miles + away_travel_high + attendance, data = games)

lm_model_4 <- lm(point_diff ~ dist_miles + away_travel_high + attendance + year, data = games)
```

### **Model Specification**

Given the large sample size, we use a linear regression model to examine the relationship between Point Differential and predictors such as Distance Traveled, Travel Direction, Attendance, and Year. The regression table below shows four models that progressively add more predictors.

1.  Model 1: The first model includes only Distance Traveled. The coefficient for this variable is statistically insignificant, suggesting that Distance Traveled has little effect on Point Differential.
2.  Model 2: In this model, we add Travel Direction (with "Neither" as the baseline). The coefficients for West to East and East to West are statistically significant, indicating that teams traveling from West to East tend to have a higher point differential, while those traveling East to West perform worse.
3.  Model 3: Game Attendance is introduced. Although it is statistically significant, the effect is small, suggesting that larger crowds slightly affect the point differential but with limited practical significance.
4.  Model 4: Finally, we add Year based on EDA findings, and it shows a statistically significant positive effect, implying that point differentials have increased over time.

```{r model, echo = FALSE, results='asis', message=FALSE, warning=FALSE}

stargazer(
  lm_model_1, lm_model_2, lm_model_3, lm_model_4,
  header = FALSE,
  no.space = TRUE, 
  column.sep.width = "0.5pt", 
  omit.stat = "f",
  font.size = "footnotesize",
  type="latex",
  title="Linear Regression Co-efficients for Point Differential Against Input Variables", 
  dep.var.labels = "Point Differential", 
  covariate.labels = c("Distance Travelled (miles)", 
                       "West to East Direction", 
                       "East to West Direction",
                       "Game Attendance",
                       "Year",
                       "Constant"), 
  digits=3)
```

### **Model Assumptions**

Since the model is a large sample linear model, the required assumptions are as follows:

#### 1. Independent & Identically Distributed Errors

The model assumes that errors are independent and identically distributed. However, games are not fully independent, as the performance of one team can impact the performance of another team, especially in a competitive context. For example, a team's performance in one game might be influenced by the strength of its opponent, which introduces a form of dependence between teams' performances across seasons. To partially address this, we include year as a numeric covariate, which captures temporal trends in point differentials and helps account for time-dependent factors.

#### 2. A Unique Best Linear Predictor Exists

Covariance matrices (appendix figure x) show finite variance and no perfect collinearity, indicating that the distribution does not have heavy tails. Additionally, no variable is a linear combination of another variable and the correlation matrices (appendix figure y) confirms that all predictor variables are sufficiently independent for OLS estimation.

### **Model Results & Interpretation**

The final model (model 4) is statistically significant by means of the f-statistic p-value being less than 2.2e-16. Within the model, the variable distance traveled, has a coefficient of -0.0002, meaning that all else equal, for each additional mile traveled by the away team, point differential decreases by .00002. The effect of distance on point differential remains not statistically significant since the p-value is greater than 0.05 and the results are not practically significant since even the longest normal flight from Miami to Portland which is about 3000 miles, only correlates to a point differential decrease of 0.6 points.

NOT FINISHED

### **Scope & Limitations**

This analysis is descriptive in nature and does not support causal claims about the effect of travel on NBA game outcomes. Games are not fully independent, as team performance may be correlated across time due to persistent team quality, injuries, and broader league dynamics. Additionally, while geodesic distance captures the physical separation between teams, it does not directly measure fatigue, rest, travel conditions, or circadian disruption, such as back-to-back games or sleep loss from changing time zones. Finally, structural changes in the NBA over the seven-decade sample period, including standardization of courts, more movement of players between teams, and travelling being less strenuous, may influence both the distribution of point differentials and their relationship with travel-related variables.

# **Conclusion**

Our analysis reveals that distance traveled by away teams does not significantly affect point differential, suggesting that travel distance also is not a major determinant of game performance. However, travel direction matters: West Coast teams traveling to the East Coast experience a measurable advantage, while East-to-West effects are weaker and less consistent. Consistent with our exploratory analysis, home-court advantage has been declining over time, with away teams winning a larger share of games in recent years. These findings provide important insights into travel and performance dynamics in the NBA, highlighting patterns that may inform scheduling, team preparation for games, and other factors, while emphasizing the limits of causal interpretation due to data and model constraints.

\newpage

# Citations

# Appendix

Link to the Dataset : https://www.kaggle.com/datasets/eoinamoore/historical-nba-data-and-player-box-scores?utm_source=chatgpt.com&select=Games.csv

```{r test_graphs, echo = FALSE, message=FALSE, warning=FALSE, fig.width = 9, fig.height=9}
lm_model_4 <- lm(point_diff ~ dist_miles + away_travel_high + attendance + year, data = games)
games$fitted_values <- predict(lm_model_4, newdata = games)

p1 <- ggplot(games, aes(x = point_diff, y = fitted_values)) +
  geom_point(alpha = 0.5, size = 1, color = "blue") +  # Raw data points
  geom_smooth(method = "lm", color = "red", se = FALSE) +  # Regression line
  labs(
    title = "Actual Point Difference vs. Predicted Point Difference",
    x = "Distance Travelled (miles)",
    y = "Actual Point Difference"
  )

p2 <- ggplot(games, aes(x = dist_miles, y = fitted_values)) +
  geom_point(alpha = 0.5, size = 1, color = "blue") +  # Raw data points
  geom_smooth(method = "lm", color = "red", se = FALSE) +  # Regression line
  labs(
    title = "Predicted Point Difference vs. Distance Travelled",
    x = "Distance Travelled (miles)",
    y = "Predicted Point Difference"
  )

p3 <- ggplot(games, aes(x = year, y = fitted_values)) +
  geom_point(alpha = 0.5, size = 1, color = "blue") +  # Raw data points
  geom_smooth(method = "lm", color = "red", se = FALSE) +  # Regression line
  labs(
    title = "Predicted Point Difference vs. Year",
    x = "Year",
    y = "Predicted Point Difference"
  )

p4 <- ggplot(games, aes(x = attendance, y = fitted_values)) +
  geom_point(alpha = 0.5, size = 1, color = "blue") +  # Raw data points
  geom_smooth(method = "lm", color = "red", se = FALSE) +  # Regression line
  labs(
    title = "Predicted Point Difference vs. Attendance",
    x = "Attendance",
    y = "Predicted Point Difference"
  )

p5 <- ggplot(games, aes(x = away_travel_high, y = fitted_values)) +
  geom_boxplot(fill = "lightblue", color = "darkblue") +
  labs(title = "Predicted Point Difference vs. Away Travel High", x = "Away Travel High", y = "Predicted Point Difference")

p1+p2+p3+p4+p5 + plot_layout(ncol = 2)
```

```{r residual_fitted, echo = FALSE, message=FALSE, warning=FALSE, fig.width = 9, fig.height=9}
games$fitted_values <- predict(lm_model_4, newdata = games)
games$residuals <- games$point_diff - games$fitted_values

ggplot(games, aes(x = fitted_values, y = residuals)) +
  geom_point(alpha = 0.5, size = 1, color = "blue") +  # Raw data points
  geom_hline(yintercept = 0, color = "red") +  # Add a horizontal line at y = 0
  xlim(-12, 2) +
  labs(
    title = "Residuals vs Predicted Values",
    x = "Predicted Values (Fitted)",
    y = "Residuals"
  ) 
```

NOTE: THIS GRAPH HAS A WHITE LINE BECAUSE 0 VALUE IS IMPOSSIBLE NBA GAMES CAN'T BE TIED AND WILL ALWAYS GO INTO OVERTIME UNTIL SOMEONE WINS

```{r,  echo = FALSE, results='asis', message=FALSE, warning=FALSE, fig.width=150}

cov_cor_cleaned_data <- na.omit(games[, c("dist_miles", "year", "attendance", "away_travel_high", "point_diff")])
cov_cor_cleaned_data$travel_direction <- as.numeric(cov_cor_cleaned_data$away_travel_high)
cov_matrix <- cov(cov_cor_cleaned_data[, c("dist_miles", "year", "attendance","travel_direction", "point_diff")])
# Format the covariance matrix for better alignment
print(cov_matrix, digits =3)

```

```{r,  echo = FALSE, results='asis', message=FALSE, warning=FALSE}
cor_matrix <- cor(cov_cor_cleaned_data[, c("dist_miles", "year", "attendance","travel_direction", "point_diff")])
print(cor_matrix,digits =3)

```
